\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-03-15 doc/incertae_sedis_latex2.tex
Update TEX and PDF.

We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
The taxonomy tree is considered by ranked below any of the input trees.
Each node $i$ in the taxonomy tree is associated a taxon name $n(i)$
in the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
and contains all taxa. In contrast, input phylogenies only have labels
on their leaf nodes, and generally include a small fraction of the
total number of taxa. Furthermore, taxa in input trees need not correspond
to leaf taxa on the taxonomy tree, and multiple tips may be labeled
with the same taxon name. \citet{redelings2017supertree} described
a method called 'exemplification' that reduces this problem to a simpler
form where all taxon names of input trees (i) correspond to leaf taxa
and (ii) occur only once in an input tree. For the purposes of this
paper, we therefore assume that this is the case.

@@

\subsection{Splits}

Each edge of a standard tree divides the tip taxa $\mathcal{L}$ into
two groups: the include set $\mathcal{I}(e)$ which does not contain
the root, and the exclude set $\mathcal{E}(e)$ which does contain
the root. Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e),
\]
with the exclude set always on the right. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

@@

\subsubsection{Partially reduced exclude sets for \emph{incertae sedis} taxa}

Marking a taxon as \emph{incertae sedis} changes the meaning of its
sibling taxa. Since the \emph{incertae sedis} taxon can be placed
inside its sibling subtrees, the exclude sets of the sibling subtrees
should not include the \emph{incertae sedis }taxon. Although exclude
sets of sibling taxa are shrunk, the include sets of the sibling taxa
are unaffected.

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude $\mathcal{E}(n)$ set for
a node $n$ contains the children of all the ancestors of $n$, unless
those children are incertae sedis. We can compute exclude sets recursively
by writing the exclude set of a node $n$ in terms of the exclude
set of its parent. If we use the terminology that the include and
exclude sets for a node $n$ are $\mathcal{I}(n)$ and $\mathcal{E}(n)$,
then we have 
\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-1}
\end{align}
This leads to a pre-order recursion that terminates if the exclude
set for the root note is set to $\{\}$ as a boundary condition.

\subsubsection{Extended formulation}

If an \emph{incertae sedis} taxon $A$ is placed within a sibling
clade, then it may gain a new \emph{incertae sedis} sibling $B$.
We could then possibly place $B$ within $A$. However, the above
definition does not allow this possibility. We therefore adjust the
above definition to exclude only descendants of sibling taxon only
the path to the descendant does not pass through a node labeled as
\emph{incertae sedis}. Such descendants of a sibling $m$ are given
by:
\begin{align*}
\mathcal{I}^{\prime}(m) & =\begin{cases}
m & \text{if }m\text{ is a leaf}\\
\left[\mathcal{I}^{\prime}(c)\big|c\in children(m),c\text{ not \emph{incertae sedis}}\right] & \text{otherwise.}
\end{cases}
\end{align*}
We then modify the formula above to exclude $\mathcal{I}^{\prime}(m)$
instead of $\mathcal{I}(m)$ for each sibling $m$:
\begin{align}
\mathcal{E}^{\prime}(n) & =\mathcal{E}^{\prime}(parent(n))\cup\left[\mathcal{I}^{\prime}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-2}
\end{align}

This less stringent formula is the one we will apply throughout this
paper.

An definition in terms of moving node $A$ to position $x$ could
cause trouble when moving multiple nodes, since the order of the movement
would matter.

@@

When \emph{incertae sedis} clades are allowed, it is possible for
an \emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon. In such a case we must allow the higher-ranked taxon
to be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure \#)

@@

@@

>>>>

We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
The taxonomy tree is ranked below any of the input trees. Each node
$i$ in the taxonomy tree is associated a taxon name $n(i)$ in the
set of taxon names $\mathcal{N}$, and each taxon name is corresponds
to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
and contains all taxa. In contrast, input phylogenies only have labels
on their leaf nodes, and generally include a small fraction of the
total number of taxa. Furthermore, taxa in input trees may correspond
to higher taxa on the taxonomy tree, and multiple tips may be labeled
with the same taxon name. \citet{redelings2017supertree} described
a method called 'exemplification' that reduces this problem to a simpler
form where all taxon names of input trees (i) correspond to leaf taxa
and (ii) occur only once in an input tree. For the purposes of this
paper, we therefore assume that this is the case.

@@

\subsection{A splits-based semantics for \emph{incertae sedis} taxa}

\subsubsection{Splits}

Each edge of a standard tree divides the tip taxa $\mathcal{L}$ into
two groups: the include set $\mathcal{I}(e)$ which does not contain
the root, and the exclude set $\mathcal{E}(e)$ which does contain
the root. Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e).
\]
where the $\bullet$ indicates the root. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

@@

\subsubsection{Reduced exclude sets allow \emph{incertae sedis} to be placed into
sibling taxa}

Marking a taxon as \emph{incertae sedis} changes the meaning of its
sibling taxa. We seek to represent this by constructing modified splits
for each branch of the taxonomy tree. The include sets of these splits
remain unchanged, but we construct reduced exclude sets to indicate
cases where \emph{incertae sedis} taxa are allowed to intrude into
a taxon.

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude set $\mathcal{E}(n)$ for
node $n$ should not include any of its siblings marked \emph{incertae
}sedis\emph{,} but should exclude the descendants of its other siblings\emph{.}
Additionally, the exclude $\mathcal{E}(n)$ should contain the children
of all the ancestors of $n$, unless those children are incertae sedis.

We may also consider what happens if an \emph{incertae sedis} taxon
$A$ is placed within a sister group. Suppose $A$ is placed next
to a sibling $B$. If $B$ is \emph{incertae sedis}, then it should
be possible to place $B$ within $A$. As a result, $A$ should not
exclude descendants of $B$ than can only be reached from $A$ by
passing through a node marked \emph{incertae sedis. }

We can compute exclude sets recursively by writing the exclude set
of a node $n$ in terms of the exclude set of its parent:
\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I^{\prime}}(m)\big|m\in siblings(n)\right],\label{eq:exclude-set-formula-1}
\end{align}
where $\mathcal{I}^{\prime}(n)$ indicates the descendants of $n$
that are reachable from $n$ without passing through an \emph{incertae
sedis} node:
\begin{align*}
\mathcal{I}^{\prime}(n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}}\\
\{n\} & \text{if }n\text{ is a leaf node}\\
\left[\mathcal{I^{\prime}}(m)\big|m\in children(n)\right] & \text{if }n\text{ is an internal node}
\end{cases}
\end{align*}
This recursion terminates if the exclude set for the root node is
set to $\emptyset$ as a boundary condition.

@@

When \emph{incertae sedis} clades are allowed, it is possible for
an \emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon. In such a case we must allow the higher-ranked taxon
to be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure \ref{fig:Nested-incertae-sedis}).

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}

}\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}

\subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}

}\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}

\caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa. (a)
Input tree \#1 places CI1 within CG, and C within $A$. (b) In the
taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
sedis} under the root. (c) The synthesis tree when \emph{incertae
sedis} taxa are not considered. The names AB, A, and CG are lost.
(d) The synthesis tree when \emph{incertae sedis} taxa are considered.
The names AB, A, and CG are regained, as well as the edges leading
to AB and CG.}

@@

\emph{This case makes placement more complicated. Use a sequence of
ordered placements, preorder?}
\begin{figure}

\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}

}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}

}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}

\caption{Placing one incertae sedis group within another. Here I is placed
within A, and then AI is placed within A.}

@@

The splits-based semantics for incertae sedis taxa has many desirable
properties. In allows the use of the BUILD algorithm to assess compatibility
of split sets. However, it also has some properties that may not be
expected. For example, two incertae sedis siblings can be freely interdigitated,
which might not be the expected outcome. The current definition, also
ignore ranks. Thus if we had two incertae sedis siblings that were
genera, but other siblings were families, then one might expect that
the genus-level incertae sedis taxa could be placed within families,
but not intermixed. Since the current approach ignores ranks, it cannot
do that.

The above definition of incertae sedis taxa does allow two incertae
sedis sister taxa to be interdigitated. It also
